---
title: "IS 6813 EDA"
author: "Hunter Nilsen"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: yes
    highlight: espresso
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load the Data

```{r}
# Load the CSV file
data <- read.csv("IWC_Work_Orders_Extract.csv", stringsAsFactors = FALSE)
```

## Load Packages

```{r}
library(dplyr)
library(tidyverse)
library(VIM)
library(corrplot)
library(GGally)
library(factoextra)
```

## Explore the Data

```{r}
# View the first few rows of the dataset
head(data)

# Get an overview of the dataset
str(data)

# Summary statistics for numeric columns
summary(data)
```

This step gives an initial glimpse of the dataset, including the first few rows, data types, and summary statistics. This helps understand the data distribution and identify potential missing or erroneous values.

## Data Cleaning

```{r}
# Convert the columns to Date format
data$EXECUTION_START_DATE <- as.Date(data$EXECUTION_START_DATE, format="%m/%d/%Y")
data$EXECUTION_FINISH_DATE <- as.Date(data$EXECUTION_FINISH_DATE, format="%m/%d/%Y")
```

Dates are converted to the correct Date format for proper handling in time-series analysis and other date-related operations.

## Summaries and Plots

```{r}
# Histogram of actual work in minutes
hist(data$ACTUAL_WORK_IN_MINUTES, main="Distribution of Downtime", xlab="Downtime (minutes)", breaks=20)
```

The histogram provides a visualization of downtime distribution. This helps identify skewness, common downtime ranges, and outliers.

```{r}
# Boxplot of downtime by plant
boxplot(ACTUAL_WORK_IN_MINUTES ~ PLANT_ID, data=data, main="Downtime by Plant", xlab="Plant", ylab="Downtime (minutes)")
```

The boxplot compares downtime distributions across different plants, highlighting variability and potential outliers in specific locations.

```{r}
# Barplot for the frequency of Maintenance Activity Types (planned vs unplanned)
barplot(table(data$MAINTENANCE_ACTIVITY_TYPE), main="Maintenance Activity Type", ylab="Frequency", xlab="Type")
```

The barplot shows the frequency of planned versus unplanned maintenance activities, which can inform maintenance planning strategies.

## Missing Data

```{r}
# Visualize missing data patterns
aggr(data, col=c('navyblue','red'), numbers=TRUE, sortVars=TRUE, labels=names(data), cex.axis=.7, gap=3, ylab=c("Missing data","Pattern"))
```

This visualization highlights patterns of missing data, aiding decisions on data imputation or removal of columns/rows.

```{r}
# Calculate percentage of missing data per column
missing_percent <- colSums(is.na(data)) / nrow(data) * 100
print(missing_percent)
```

The percentage of missing data per column is calculated to understand the extent of missingness and decide on the appropriate treatment.

## Correllation Analysis

```{r}
# Subset numeric columns
numeric_data <- data[, sapply(data, is.numeric)]

# Correlation matrix
cor_matrix <- cor(numeric_data, use="complete.obs")
print(cor_matrix)

# Visualize the correlation matrix using corrplot

corrplot(cor_matrix, method = "circle")
```

The correlation matrix and its visualization help identify relationships between numeric variables. Strong correlations might indicate redundant variables or potential predictors.

## Categorical Variable Analysis

```{r}
# Summarize downtime by Maintenance Type
downtime_by_type <- aggregate(ACTUAL_WORK_IN_MINUTES ~ MAINTENANCE_TYPE_DESCRIPTION, data, sum)

# Barplot of total downtime by maintenance type
barplot(downtime_by_type$ACTUAL_WORK_IN_MINUTES, names.arg = downtime_by_type$MAINTENANCE_TYPE_DESCRIPTION, las=2, main="Downtime by Maintenance Type", ylab="Downtime (Minutes)")
```

The total downtime is summarized by maintenance type, providing insights into which types of maintenance consume the most time.

```{r}
# Boxplot of downtime for different maintenance types
ggplot(data, aes(x=MAINTENANCE_TYPE_DESCRIPTION, y=ACTUAL_WORK_IN_MINUTES)) +
  geom_boxplot() +
  labs(title="Downtime by Maintenance Type", x="Maintenance Type", y="Downtime (Minutes)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

This boxplot highlights downtime variability within each maintenance type, showing the spread and potential outliers.

## Outlier Detection

```{r}
# Boxplot to detect outliers in downtime
boxplot(data$ACTUAL_WORK_IN_MINUTES, main="Boxplot for Downtime", ylab="Downtime (Minutes)")
```

The boxplot provides a simple way to identify outliers in downtime data.

```{r}
# Detect outliers using Z-scores
z_scores <- scale(data$ACTUAL_WORK_IN_MINUTES)
outliers <- which(abs(z_scores) > 3)
print(outliers)
```

Outliers are detected using Z-scores, which highlight extreme values beyond three standard deviations from the mean.

```{r}
# Detect outliers using Z-scores
z_scores <- scale(data$ACTUAL_WORK_IN_MINUTES)
outliers <- which(abs(z_scores) > 3)
print(outliers)
```

The pairwise scatter plot matrix visualizes relationships between numeric variables, providing insights into linear or non-linear patterns.

## Pairwise Relationships

```{r}
# Pairwise scatter plot matrix


ggpairs(numeric_data)
```

## Feature Engineering

```{r}
# Example: Create a new feature "Downtime Category"
data$downtime_category <- cut(data$ACTUAL_WORK_IN_MINUTES, breaks=c(0, 60, 120, 180, Inf), labels=c("Short", "Medium", "Long", "Very Long"))

# Check the distribution of this new feature
table(data$downtime_category)
```

A new categorical feature downtime_category is created to group downtime into meaningful categories, enabling more targeted analysis.

## Cross Tabulation

```{r}
# Cross-tabulation of maintenance type and functional location
table(data$MAINTENANCE_TYPE_DESCRIPTION, data$FUNCTIONAL_LOC)
```

This cross-tabulation shows the distribution of maintenance types across different functional locations, revealing patterns or dependencies.

## Heatmaps

```{r}
# Install and load necessary packages
install.packages("ggplot2")
install.packages("reshape2")

# Create a heatmap
library(ggplot2)
library(reshape2)
cor_data <- cor(numeric_data, use = "complete.obs")
melted_cor_data <- melt(cor_data)

ggplot(data = melted_cor_data, aes(x=Var1, y=Var2, fill=value)) + 
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, limit = c(-1,1), space = "Lab", name="Correlation") +
  theme_minimal() +
  labs(title="Correlation Heatmap", x="", y="")
```

The heatmap provides an intuitive visualization of the correlation matrix, showing relationships between variables at a glance.

## Correlations

```{r}
# Subset numeric columns
numeric_data <- data[, sapply(data, is.numeric)]

# Check the first few rows of the numeric data
head(numeric_data)
```

```{r}
# Compute correlation matrix
cor_matrix <- cor(numeric_data, use="complete.obs")
print(cor_matrix)
```

